### 2019-07-24 reports/thesis/chapters/3_compoundPoisson.tex
start on compound Poisson chapter

In this chapter, the compound Poisson-Gamma distribution is reviewed and studied. Firstly methods for evaluating the density of the compound Poisson-Gamma distribution are discussed. Approximations such as the Normal approximation or saddlepoint approximation \citep{daniels1954saddlepoint} can be used. These approximate densities can be written in closed form. Exact methods can be computationally challenging when dealing with the infinite sum but can be done when summing over important terms \citep{dunn2005series}. Next, a method for parameter estimation will be proposed here using the EM algorithm \citep{dempster1977maximum}. However in the end of the chapter, it will be shown that no unique maximum likelihood estimator exist when certain parameters of the compound Poisson-Gamma distribution gets too large.

\section{Density Evaluation}
Suppose $X\sim\CPoisson(\lambda,\alpha,\beta)$ and $\lambda,\alpha,\beta>0$ then the p.d.f.~of $X$ is
\begin{equation}
	p_X(x) = 
	\begin{cases}
		\delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
		\euler^{-\beta x-\lambda}\frac{1}{x}\sum_{y=1}^{\infty}W_y & \text{ for } x>0 
	\end{cases}
\end{equation}
where
\begin{equation}
	W_y = \frac{\beta^{y\alpha}\lambda^yx^{y\alpha}}{\Gamma(y\alpha)y!}
	\ .
	\label{eq:compoundPoisson_w}
\end{equation}
The evaluation of this p.d.f.~is difficult because the sum of $W_y$ cannot be simplified. However the m.g.f.~of $X$ can be written down in closed form
\begin{equation}
	M(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
	\ .
\end{equation}
The m.g.f.~allows the study of limiting distributions of the compound Poisson-Gamma distribution. These limiting distributions then can be used to approximate the p.d.f.~of a compound Poisson-Gamma random variable.

The saddlepoint approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation  of the m.g.f.~giving the p.d.f. However inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddlepoint approximation will be studied here.

Lastly the infinite sum can be computationally summed in a clever way. This was done by summing only large terms in the sum and ignores small terms. By using Stirling's approximation, the largest term in the sum can be approximately found \citep{dunn2005series}.

\subsection{Normal Approxmation}
The m.g.f.~provides a starting point to what limiting distributions the compound Poisson-Gamma distribution converges to for large parameters. However the m.g.f.~of $X$ itself it not useful because when considering $\lambda\rightarrow\infty$ or $\alpha\rightarrow\infty$ then $M_X(\theta)\rightarrow\infty$. Also for $\beta\rightarrow\infty$, then $M_X(\theta)\rightarrow 1$ which is not useful either.

The compound Poisson-Gamma random variable $X$ should be standardised to obtain useful limiting results from the m.g.f. Let
\begin{equation}
	Z = \frac{X-\expectation[X]}{\sqrt{\variance[X]}}
	\ ,
\end{equation}
then using previous results
\begin{equation*}
	Z = \frac{\beta}{\sqrt{\alpha(\alpha+1)\lambda}}
	\left(
		X-\frac{\alpha\lambda}{\beta}
	\right)
\end{equation*}
\begin{equation}
	Z = bX+a
\end{equation}
where
\begin{equation}
	b = \frac{\beta}{\sqrt{\alpha(\alpha+1)\lambda}}
\end{equation}
and
\begin{equation}
	a = -\sqrt{\frac{\alpha\lambda}{\alpha+1}}
	\ .
\end{equation}

The m.g.f.~of $Z$ is then
\begin{equation*}
	M_Z(\theta)=\expectation\left[\euler^{Z\theta}\right]
\end{equation*}
\begin{equation*}
	M_Z(\theta)=\expectation\left[\euler^{(bX+a)\theta}\right]
\end{equation*}
\begin{equation*}
	M_Z(\theta)=\euler^{a\theta}M_X(b\theta)
	\ .
\end{equation*}
Substituting in the values
\begin{equation*}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\left(
			\left(
				\frac{\beta}{\beta-\frac{\beta\theta}{\sqrt{\alpha(\alpha+1)\lambda}}}
			\right)^\alpha
			-1
		\right)
	\right]
\end{equation*}
\begin{equation*}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\left(
			\left(
				\frac{\sqrt{\alpha(\alpha+1)\lambda}}{\sqrt{\alpha(\alpha+1)\lambda}-\theta}
			\right)^\alpha
			-1
		\right)
	\right]
\end{equation*}
or in a different form
\begin{equation}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\left(
			\left(
				1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}	
			\right)^{-\alpha}
			-1
		\right)
	\right]
	\ .
\end{equation}

The above form is useful so that the binomial expansion can be conducted, that is
\begin{multline}
	\left(
		1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}
	\right)^{-\alpha}
	=
	1+
	\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\\
	\text{for }\frac{\|\theta\|}{\sqrt{\alpha(\alpha+1)\lambda}}<1 \ .
\end{multline}
Substituting in the binomial expansion
\begin{equation*}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\right]
	\ .
\end{equation*}
Writing in full the $r=1,2$ terms
\begin{multline*}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
		+\frac{\theta^2}{2}
	\right.
	\\
	\left.
		+\lambda\sum_{r=3}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\right]
\end{multline*}
and a term cancels out to get
\begin{equation}
	M_Z(\theta)=
	\exp\left[
		\frac{\theta^2}{2}
		+\sum_{r=3}^\infty
		\frac
			{\theta^r\prod_{s=1}^r(\alpha+s-1)}
			{(\alpha(\alpha+1))^{r/2}r!}
		\lambda^{1-r/2}
	\right]
	\ .
\end{equation}

For large $\lambda$
\begin{equation}
	\lim_{\lambda\rightarrow\infty}M_Z(\theta)=\exp\left[\frac{\theta^2}{2}\right]
\end{equation}
which is the same as the m.g.f.~of a standard Normal variable, therefore
\begin{equation}
	\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
This should make sense as for high $\lambda$, the Poisson random variable has a high expectation, increasing the number of Gamma random variables in a summation. Increasing the number of terms in a summation will trigger the central limit theorem.

Next for high $\alpha$, it should be noted that
\begin{equation*}
	\lim_{\alpha\rightarrow\infty}
	\frac{
		\prod_{s=1}^r(\alpha+s-1)
	}
	{
		\alpha(\alpha+1))^{r/2}
	}
	=
	\frac{
		\prod_{s=1}^r\alpha
	}
	{
		\alpha^{r}
	}
\end{equation*}
so that
\begin{equation}
	\lim_{\alpha\rightarrow\infty}
	\frac{
		\prod_{s=1}^r(\alpha+s-1)
	}
	{
		(\alpha(\alpha+1))^{r/2}
	}
	= 1
	\ .
\end{equation}
As a result
\begin{equation}
	\lim_{\alpha\rightarrow\infty}
	M_Z(\theta)=
	\exp\left[
		\frac{\theta^2}{2}
		+\sum_{r=3}^\infty
		\frac{\theta^r}{r!}
		\lambda^{1-r/2}
	\right]
	\ .
\end{equation}
However this shows that taking the limit $\alpha\rightarrow\infty$ is not enough to get a Normal limiting distribution. For a Normal limiting distribution, the limit must be accompanied with the limit $\lambda\rightarrow\infty$, that is
\begin{equation}
	\lim_{\lambda\rightarrow\infty}\lim_{\alpha\rightarrow\infty}Z\sim\normal(0,1)
	\ .
\end{equation}

@@

\subsection{Normal Approximation - General Case}
It was shown in certain situations, the compound Poisson-Gamma distribution can be approximated with the Normal distribution. This can be extended to the general compound Poisson.

Let $Y\sim\poisson(\lambda)$ where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some random variable for i.i.d.~$i=1,2,3,\dotdotdot$ with m.g.f.~$M_U(\theta)$. Let
\begin{equation}
	X|Y = \sum_{i=1}^{Y}U_i
	\ .
\end{equation}

>>>>

The grey value of each pixel in the detector can be modelled as a random variable, due to the random behaviour of photons being produced, interacting with the test sample and the scintillator in the detector. By modelling using a random variable, the uncertainty can be quantified and considered when conducting inference about any detected defects.

The compound Poisson distribution is studied here because of the compound Poission-like behaviour from the detection of photons. It is defined by defining a latent variable $Y\sim\poisson(\lambda)$ with probability mass function (p.m.f.)
\begin{equation}
  \prob(Y=y)=\euler^{-\lambda}\frac{\lambda^y}{y!} \quad \text{for }y=0,1,2,\cdots
\end{equation}
where $\lambda>0$ is the Poisson rate parameter. Let $U_i$ be some independent and identically distributed (i.i.d.) latent random variables with probability density function (p.d.f.) $p_U(u)$ for $i=1,2,3,\dotdotdot$. Let $X$ be a compound Poisson random variable where
\begin{equation}
  X|Y = \sum_{i=1}^{Y}U_i \ .
  \label{eq:compoundPoisson_X|Y}
\end{equation}
The p.d.f.~of $X$ can be obtained by marginalising the joint p.d.f.
\begin{equation}
  p_X(x)=\sum_{y=0}^\infty p_{X|Y}(x|y)\prob(Y=y) \quad\text{for }x\geqslant 0
  \ .
\end{equation}
It should be noted that $X=0$ if and only if $Y=0$ with probability $\prob(X=0)=\euler^{-\lambda}$. Then $X$ has probability mass at $X=0$ and probability density at $X>0$ which results in the p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda}  & \text{ for } x=0 \\ 
    \sum_{y=1}^\infty p_{X|Y}(x|y)\euler^{-\lambda}\frac{\lambda^y}{y!} \quad\text{for } & \text{ for } x>0
  \end{cases}
\end{equation}
where $\delta(x)$ is the Dirac delta function.

This chapter starts with a literature review on the compound Poisson distribution, how it is derived from the behaviour of photons, how its likelihood is evaluated and methods for fitting onto data. A model was proposed for the pixel's grey values and the expectation-maximization (EM) algorithm was implemented to fit the model onto data. It was found that for high photon rate, there were identifiability issues. The chapter is concluded on a discussion on why the EM algorithm failed.

\section{Literature Review}

\subsection{Compound Poisson in x-ray detection}

In an x-ray tube, photons are emitted as a Poisson process and each photon has some random energy due to bremsstrahlung and characteristic radiation. This shares similarities to the compound Poisson. Let $Y$ be the number of photons emitted for some time exposure $\tau$, then $Y\sim\poisson(\lambda)$. Each photon is assumed to be i.i.d.~with energy $U_i$ for $i=1,2,3,\cdots$. $U_i$ has p.d.f.~$p_U(u)$. The random variables discussed here covers all the latent variables in the compound Poisson.

Photons emitted from the x-ray tube undergo attenuation when propagating through the test sample. Assuming no beam hardening, some photons are either absorbed or scattered, making them undetectable. Scattered photons may be detected but it is very rare. Attenuated photon energy remains unaffected so attenuation decreases the parameter $\lambda$ and the amount it decreases by depends on the attenuation coefficient of the material and the amount of material the x-ray attenuates. The parameters of the random variable $U_i$ remains unchanged because the energy of each photon remains the same after attenuation, assuming no beam hardening.

When the photons interact with the scintillator in the detector, they are converted into visible light. The visible light photons are then detected and converted into a a grey value. A quantum counter set the digital signal to be linear with to the number of photons detected \citep{whiting2006properties}. Let $X$ be the grey value observed, then
\begin{equation}
X = bY + \epsilon
\end{equation}
where $\epsilon\sim\normal(a,\kappa)$, $b$ and $a$ are some constant and $\kappa$ is the variance of electronic noise. In a quantum counter, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = b\lambda + a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = b^2\lambda + \kappa
\end{equation}
respectively. By eliminating $\lambda$
\begin{equation}
\variance\left[X\right] = b\expectation\left[X\right]+\kappa-ab \ ,
\end{equation}
a linear relationship between the variance and expectation of the grey value is obtained.

An energy integrating detector records the grey value which is linear to the energy detected \citep{whiting2006properties}. The grey value $X$ is
\begin{equation}
X|Y = \sum_{i=1}^Y U_i + \epsilon \ .
\end{equation}
This is the compound Poisson with Normal noise added to it. The scale factor $b$ is not included as this can be absorbed into $U$. Using the result that $\expectation\left[X\right]=\expectation\expectation\left[X|Y\right]$ and $\variance\left[X\right] = \variance\expectation\left[X|Y\right] + \expectation\variance\left[X|Y\right]$, the mean and variance of the grey value are
\begin{equation}
\expectation\left[X\right] = \lambda\expectation\left[U\right]+a
\end{equation}
and
\begin{equation}
\variance\left[X\right] = \lambda \expectation\left[U^2\right]+\kappa
\end{equation}
respectively. Eliminating $\lambda$ obtains
\begin{equation}
\variance\left[X\right] = \dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \expectation\left[X\right] + \kappa - a\dfrac{\expectation\left[U^2\right]}{\expectation\left[U\right]} \ .
\end{equation}
By assuming no beam hardening, $\expectation\left[U\right]$ and $\expectation\left[U^2\right]$ remains constant, thus there is a linear relationship between the variance and expectation of the grey value. There are other types of detection schemes \citep{whiting2006properties} but it shall be not be considered here.

Experiments have been done to verify the compound Poisson nature of the detector. This was done by investigating the variance of radiographs of air \citep{hsieh2015compound} and a polyethylene cylinder \citep{yang2009evaluation} \citep{yang2010noise} at different voltages and powers. It was found there were 2 components in the noise, one was signal dependent and comes from the compound Poisson, the other was signal independent and may be electronic noise. The electronic noise can be modelled as Normally distributed \citep{xu2009electronic}.

\subsection{Moment generating function}

Returning back to the compound Poisson with no electronic noise $X|Y = \sum_{i=1}^{Y}U_i$. Let the moment generating function (m.g.f.) of $X$ be
\begin{equation}
  M_X(\theta)=\expectation\left[\euler^{X\theta}\right]
  \ .
\end{equation}
This can be computed using the result for conditional expectations $M_X(\theta)=\expectation\expectation\left[\euler^{X\theta}|Y\right]$. Using the definition of $X|Y$ in Equation \eqref{eq:compoundPoisson_X|Y}, then
\begin{align}
  M_X(\theta)&=\expectation\expectation\left[\exp\left(\theta U_1 + \theta U_2 + \dotdotdot + \theta U_Y\right)|Y\right]
  \\
  &=\expectation\expectation\left[\euler^{\theta U_1}\cdot\euler^{\theta U_2}\cdot\dotdotdot\cdot\euler^{\theta U_Y}|Y\right]
  \ .
\end{align}
But because $U_i$ for $i=1,2,3,\dotdotdot$ are i.i.d., then each $U_i$ has a common m.g.f.~$M_U(\theta)=\expectation\left[\euler^{U\theta}\right]$, then
\begin{align}
  M_X(\theta)&=\expectation\left(
    \expectation\left[\euler^{\theta U_1}|Y\right]\cdot
    \expectation\left[\euler^{\theta U_2}|Y\right]\cdot
    \dotdotdot \cdot
    \expectation\left[\euler^{\theta U_Y}|Y\right]
  \right)
  \\
  &=\expectation\left[\left(M_U(\theta)\right)^Y\right]
  \\
  &=\expectation\left[\euler^{Y\ln(M_U(\theta))}\right]
  \\
  & = M_Y\left(\ln(M_U(\theta)\right)
\end{align}
where $M_Y(\theta)$ is the m.g.f.~of $Y$. It can be shown that the m.g.f.~of $Y$ is
$
  M_Y(\theta)=\expectation\left[\euler^{Y\theta}\right]=
  \exp
  \left[
    \lambda
    \left(
      \euler^\theta-1
    \right)
  \right]
$
then
\begin{equation}
  M_X(\theta)=
  \exp\left[
    \lambda
    \left(
      M_U(\theta)-1
    \right)
  \right]
  \ .
\end{equation}

Moments of $X$ can be obtained from the m.g.f.~by differentiating it and setting it to zero
\begin{align}
  M_X'(\theta)&=\exp\left[\lambda\left(M_U(\theta)-1\right)\right]\cdot\lambda M_U'(\theta) \\
  &=M_X(\theta)\lambda M_U'(\theta)
\end{align}
then
\begin{equation}
  \expectation\left[X\right]=\lambda\expectation\left[U\right]
  \ .
\end{equation}
Conducting the same procedure
\begin{equation}
  M_X''(\theta)=M_X'(\theta)\lambda M_U'(\theta)+M_X(\theta)\lambda M_U''(\theta)
  \ ,
\end{equation}
the variance can be obtained
\begin{align}
  \variance\left[X\right]&=M_X''(0)-\left[M_X'(0)\right]^2
  \\
  &=M_X'(0)\lambda M_U'(0)+M_X(0)\lambda M_U''(0)-\left[\lambda\expectation\left[U\right]\right]^2
  \\
  &=\lambda^2 \left(\expectation\left[U\right]\right)^2+\lambda \expectation\left[U^2\right]-\left[\lambda\expectation\left[U\right]\right]^2
\end{align}
resulting in
\begin{equation}
  \variance\left[X\right] = \lambda\expectation\left[U^2\right]
  \ .
\end{equation}
This can be extended for higher moments.

Because the m.g.f.~can be written down in closed form, so is the characterisation function or Fourier transform of the p.d.f.~of $X$. By obtaining an empirical version of $p_U(u)$, moments of $X$ can be estimated using the m.g.f.~and the p.d.f.~can be estimated by using fast Fourier transform on the empirical characteristic function \citep{whiting2006properties}.

\subsection{Generalised linear model}

A special case of the compound Poisson is when $U\sim\gammaDist\left(\alpha,\beta\right)$ where $\alpha>0$ is the gamma shape parameter and $\beta>0$ is the gamma rate parameter. The special case is used for example in \cite{xu2009electronic}. This is known as the compound Poisson-gamma distribution and denoted as $X\sim\CPoisson(\lambda,\alpha,\beta)$ and has p.d.f.
\begin{equation}
  p_X(x) = 
  \begin{cases}
    \delta(x) \euler^{-\lambda} & \text{ for } x=0 \\ 
    \sum_{y=1}^{\infty}\frac{\beta^{y\alpha}}{\Gamma(y\alpha)}x^{y\alpha-1}\euler^{-\beta x}\euler^{-\lambda}\frac{\lambda^y}{y!} & \text{ for } x>0
  \end{cases}
  \ .
  \label{eq:compoundPoisson_pdf}
\end{equation}
The conditional distribution can be shown to be $X|Y\sim\gammaDist\left(Y\alpha,\beta\right)$.

It can be shown that the compound Poisson-gamma distribution is in the exponential family for fixed $\alpha$ \citep{jorgensen1987exponential}. To show this, it is useful to reparametrize the compound Poisson-gamma distribution using the following:
\begin{equation}
  p=\frac{2+\alpha}{1+\alpha}
  \ ,
\end{equation}
\begin{equation}
  \mu=\frac{\lambda\alpha}{\beta}
  \ ,
\end{equation}
\begin{equation}
  \phi = \frac{\alpha+1}{\beta^{2-p}(\lambda\alpha)^{p-1}}
  \ .
\end{equation}
The parameters $p$, $\mu$ and $\phi$ are called the index, mean and dispersion parameters respectively. It can be shown that $1<p<2$, $\mu>0$ and $\phi>0$.

By rearranging the parameters to get
\begin{equation}
  \lambda=\frac{\mu^{2-p}}{\phi(2-p)}
\end{equation}
\begin{equation}
  \alpha=\frac{2-p}{p-1}
\end{equation}
\begin{equation}
  \beta=\frac{1}{\phi(p-1)\mu^{p-1}}
\end{equation}
and substituting it into Equation \eqref{eq:compoundPoisson_pdf}, the p.m.f.~at zero can be shown to be
\begin{equation}
  \prob(X=0) = \exp
  \left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
\end{equation}
and the p.d.f.~for $x>0$ is
\begin{multline}
  p_X(x) = \sum_{y=1}^{\infty}
  \left[
    \frac{1}{\phi(p-1)\mu^{p-1}}
  \right]^{y\alpha}
  \frac{1}{\Gamma(y\alpha)}
  x^{y\alpha-1}
  \exp\left[
      -\frac{x}{\phi(p-1)\mu^{p-1}}
  \right]
  \\
  \exp\left[
      -\frac{\mu^{2-p}}{\phi(2-p)}
  \right]
  \left[
    \frac{\mu^{2-p}}{\phi(2-p)}
  \right]^y
  \frac{1}{y!}
  \ .
\end{multline}
Tidying up the equation
\begin{multline}
  p_X(x) = 
  \exp\left[
    \frac{1}{\phi}\left(x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}\right)
  \right]
  \frac{1}{x}
  \\
  \sum_{y=1}^{\infty}\frac{x^{y\alpha}\mu^{y[2-p-\alpha(p-1)]}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{multline}
To simplify further, it should be noted that $2-p-\alpha(p-1) = 2-p - \frac{2-p}{p-1}(p-1) =0$ so that
\begin{equation}
  p_X(x) = 
  \exp\left[
    \frac{1}{\phi}
    \left(
      x\frac{\mu^{1-p}}{1-p}-\frac{\mu^{2-p}}{2-p}
    \right)
  \right]
  \frac{1}{x}
  \sum_{y=1}^{\infty}W_y(x,p,\phi)
\end{equation}
where
\begin{equation}
  W_y(x,p,\phi)=\frac{x^{y\alpha}}{\phi^{y(1+\alpha)}(p-1)^{y\alpha}(2-p)^yy!\Gamma(y\alpha)}
  \ .
\end{equation}
This is in the form of a generalised linear model \citep{nelder1972generalized, nelder1972generalized_2, mccullagh1984generalized} for fixed $p$ or $\alpha$ because the above is in the form of a distribution in the dispersive exponential family. Parameter estimation then can be done via the generalised linear model framework and can be extended to include linear mixed models \citep{zhang2013likelihood}. These has applications in for example insurance claim data \citep{jorgensen1994fitting, smyth2002fitting}.

Estimating $p$ is difficult and various methods were discussed by \cite{zhang2013likelihood}. One easy way is to estimate $\mu$ and $\phi$ on a grid of $p$'s and then select the $p$ which maximises the likelihood \citep{dunn2005series}.

One special property of the compound Poisson-gamma distribution is that it is in the Tweedie dispersion exponential family \citep{jorgensen1987exponential}. It has a special variance mean relationship
\begin{equation}
  \variance[X] = \phi \mu^p
\end{equation}
where, as a reminder, $1<p<2$. This can be derived using the partition function $Z$. Let
\begin{equation}
  \theta = \frac{\mu^{1-p}}{1-p}
\end{equation}
and
\begin{equation}
  \ln Z = \frac{1}{\phi(2-p)\theta^\alpha(1-p)^\alpha} \ ,
\end{equation}
then
\begin{equation}
  \variance[X] = \phi^2 \frac{\partial^2\ln Z}{\partial\theta^2} \ .
\end{equation}

\subsection{Method of moments}

The method of moments is a simpler method to estimate the parameters of a compound Poisson-gamma random variable. The m.g.f.~of the of $X$ is
\begin{equation}
  M_X(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
\end{equation}
and moments can be obtained from it such as
\begin{equation}
  \expectation\left[X\right]=\frac{\alpha\lambda}{\beta}
\end{equation}
\begin{equation}
  \variance\left[X\right]=\frac{\alpha(\alpha+1)\lambda}{\beta^2}
  \label{eq:compoundPoisson_variance}
\end{equation}
and
\begin{equation}
  \expectation\left[(X-\expectation[X])^3\right] = \frac{\alpha(\alpha+1)(\alpha+2)\lambda}{\beta^3}
  \ .
\end{equation}

Suppose $\widehat{\mu}$ is an estimator of $\expectation[X]$ and $\widehat{\mu}_j$ is an estimator of $\expectation\left[\left(X-\expectation[X]\right)^j\right]$ for $j=2,3$. Then the estimators
\begin{equation}
  \widehat{\lambda}=\frac{\widehat{\mu}^2\widehat{\mu}_2}{\left(2\widehat{\mu}_2^2-\widehat{\mu}_3\widehat{\mu}\right)}
\end{equation}
\begin{equation}
  \widehat{\alpha}=\frac{\widehat{\mu}_3\widehat{\mu}-2\widehat{\mu}_2^2}{\widehat{\mu}_2^2-\widehat{\mu}\widehat{\mu}_3}
\end{equation}
\begin{equation}
  \widehat{\beta}=\frac{\widehat{\mu}\widehat{\mu}_2}{\widehat{\mu}\widehat{\mu}_3-\widehat{\mu}_2^2}
\end{equation}
are method of moments estimators of $\lambda$, $\alpha$ and $\beta$ respectively \citep{withers2011compound}. These estimators suffer because estimation is not done through the sufficient statistics and can be negative, this is a problem because the parameters do not take non-positive values.

\subsection{Normal approximation}

The evaluation of the density of a compound Poisson-gamma distribution is useful so that the likelihood can be obtained. The likelihood then can be used to find, for example, maximum likelihood estimators. A problem occurs when dealing with the infinite sum in the p.d.f.~because it cannot be simplified. There are a number of approximations or computational methods to evaluate the p.d.f.~such as Fourier inverting the characteristic function \citep{dunn2008evaluation}, using the saddlepoint approximation \citep{daniels1954saddlepoint} or cleverly sum over certain terms in the infinite sum \citep{dunn2005series}. Monte Carlo methods can be used to evaluate the p.d.f.~by simulating compound Poisson-gamma random variables.

The m.g.f.~of $X$ is
\begin{equation}
M(\theta)=\exp\left[\lambda\left(\left(\frac{\beta}{\beta-\theta}\right)^{\alpha}-1\right)\right]
\ .
\end{equation}
It provides a starting point to what limiting distributions the compound Poisson-gamma distribution converges to for large parameters. These limiting distributions then can be used to approximate the p.d.f.~of a compound Poisson-Gamma random variable. The m.g.f.~of $X$ itself it not useful because when considering $\lambda\rightarrow\infty$ or $\alpha\rightarrow\infty$ then $M_X(\theta)\rightarrow\infty$. Also for $\beta\rightarrow\infty$, then $M_X(\theta)\rightarrow 1$ which is not useful either.

The compound Poisson-gamma random variable $X$ can be standardised to obtain useful limiting results from the m.g.f. Let
\begin{equation}
  Z = \frac{X-\expectation[X]}{\sqrt{\variance[X]}}
  \ ,
\end{equation}
then using previous results
\begin{equation}
  Z = bX+a
\end{equation}
where
\begin{equation}
  b = \frac{\beta}{\sqrt{\alpha(\alpha+1)\lambda}}
\end{equation}
and
\begin{equation}
  a = -\sqrt{\frac{\alpha\lambda}{\alpha+1}}
  \ .
\end{equation}

The m.g.f.~of $Z$ is then
\begin{align}
  M_Z(\theta)&=\expectation\left[\euler^{Z\theta}\right]
  \\
  &=\expectation\left[\euler^{(bX+a)\theta}\right]
  \\
  &=\euler^{a\theta}M_X(b\theta)
  \ .
\end{align}
Substituting in the values
\begin{align}
  M_Z(\theta)&=
  \exp\left(
      -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
  \exp\left[
    \lambda
    \left(
      \left(
        \frac{\beta}{\beta-\frac{\beta\theta}{\sqrt{\alpha(\alpha+1)\lambda}}}
      \right)^\alpha
      -1
    \right)
  \right]
  \\
  &=
  \exp\left(
      -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
  \exp\left[
    \lambda
    \left(
      \left(
        \frac{\sqrt{\alpha(\alpha+1)\lambda}}{\sqrt{\alpha(\alpha+1)\lambda}-\theta}
      \right)^\alpha
      -1
    \right)
  \right]
\end{align}
or in a different form
\begin{equation}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\left(
			\left(
				1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}	
			\right)^{-\alpha}
			-1
		\right)
	\right]
	\ .
\end{equation}

The above form is useful so that the binomial expansion can be conducted, that is
\begin{multline}
	\left(
		1-\frac{\theta}{\sqrt{\alpha(\alpha+1)\lambda}}
	\right)^{-\alpha}
	=
	1+
	\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\\
	\text{for }\frac{\|\theta\|}{\sqrt{\alpha(\alpha+1)\lambda}}<1 \ .
\end{multline}
Substituting in the binomial expansion
\begin{equation}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\lambda
		\sum_{r=1}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\right]
	\ .
\end{equation}
Writing in full the $r=1,2$ terms
\begin{multline}
	M_Z(\theta)=
	\exp\left(
	    -\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
    \right)
	\exp\left[
		\theta\sqrt{\frac{\alpha\lambda}{\alpha+1}}
		+\frac{\theta^2}{2}
	\right.
	\\
	\left.
		+\lambda\sum_{r=3}^\infty \frac{\theta^r\prod_{s=1}^r(\alpha+s-1)}{(\alpha(\alpha+1)\lambda)^{r/2}r!}
	\right]
\end{multline}
and a term cancels out to get
\begin{equation}
	M_Z(\theta)=
	\exp\left[
		\frac{\theta^2}{2}
		+\sum_{r=3}^\infty
		\frac
			{\theta^r\prod_{s=1}^r(\alpha+s-1)}
			{(\alpha(\alpha+1))^{r/2}r!}
		\lambda^{1-r/2}
	\right]
	\ .
\end{equation}

For large $\lambda$
\begin{equation}
	\lim_{\lambda\rightarrow\infty}M_Z(\theta)=\exp\left[\frac{\theta^2}{2}\right]
\end{equation}
which is the same as the m.g.f.~of a standard Normal variable, therefore
\begin{equation}
	\lim_{\lambda\rightarrow\infty}Z\sim\normal(0,1) \ .
\end{equation}
This should make sense as for high $\lambda$, the Poisson random variable has a high expectation, increasing the number of gamma random variables in a summation. Increasing the number of terms in a summation will trigger the central limit theorem.

Next for high $\alpha$, it should be noted that
\begin{equation}
  \lim_{\alpha\rightarrow\infty}
  \frac{
    \prod_{s=1}^r(\alpha+s-1)
  }
  {
    \alpha(\alpha+1))^{r/2}
  }
  =
  \frac{
    \prod_{s=1}^r\alpha
  }
  {
    \alpha^{r}
  }
\end{equation}
so that
\begin{equation}
  \lim_{\alpha\rightarrow\infty}
  \frac{
    \prod_{s=1}^r(\alpha+s-1)
  }
  {
    (\alpha(\alpha+1))^{r/2}
  }
  = 1
  \ .
\end{equation}
As a result
\begin{equation}
  \lim_{\alpha\rightarrow\infty}
  M_Z(\theta)=
  \exp\left[
    \frac{\theta^2}{2}
    +\sum_{r=3}^\infty
    \frac{\theta^r}{r!}
    \lambda^{1-r/2}
  \right]
  \ .
\end{equation}
However this shows that taking the limit $\alpha\rightarrow\infty$ is not enough to get a Normal limiting distribution. For a Normal limiting distribution, the limit must be accompanied with the limit $\lambda\rightarrow\infty$, that is
\begin{equation}
  \lim_{\lambda\rightarrow\infty}\lim_{\alpha\rightarrow\infty}Z\sim\normal(0,1)
  \ .
\end{equation}

@@

\subsection{Saddlepoint approximation}

The saddlepoint approximation \citep{daniels1954saddlepoint} gives an approximate solution to inverting the Laplace transformation of the m.g.f.~giving the p.d.f. Inverting the Fourier transformation of the characteristic function also gives the p.d.f.~using computational methods \citep{dunn2008evaluation}. The saddlepoint approximation will be studied here.

The m.g.f.~of a random variable $X$ is defined to be
\begin{equation}
  M_X(\theta)=\expectation[\euler^{\theta X}]
  =
  \int_{-\infty}^{\infty}\euler^{x\theta} p_X(x) \diff x
  \ .
\end{equation}
For a given m.g.f., the saddlepoint approximation \citep{daniels1954saddlepoint, butler2007saddlepoint} finds an approximate $p_X(x)$. The saddlepoint approximation is given as
\begin{equation}
  p_X(x)\approx\left(2\pi K_X''(s)\right)^{-1/2}\exp\left[K_X(s)-sx\right]
  \label{eq:saddlePoint:generalSaddlePoint}
\end{equation}
where $K_X(\theta) = \ln\left(M_X(\theta)\right)$, and $s=s(x)$ is the solution to the saddle point equation $K_X'(s)=x$.
